Embedding to Reference t-SNE Space Addresses Batch Effects in Single-Cell Classification

### 2019-08-13 paper/policar2019-tsne-embedding.tex
Shortening and diff.

Dimensionality reduction techniques, such as t-SNE, can construct informative
visualizations of high-dimensional data. When working with multiple data sets,
a straightforward application of these methods often fails; instead of
revealing underlying classes, the resulting visualizations expose
data set-specific clusters. To circumvent these batch effects, we propose an
embedding procedure that takes a t-SNE visualization constructed on a reference
data set and uses it as a scaffold for embedding new data. The new, secondary data is
embedded one data-point at the time. This prevents any interactions
between instances in the secondary data and implicitly mitigates batch
effects. We demonstrate the utility of this approach with an analysis of six
recently published single-cell gene expression data sets containing up to tens
of thousands of cells and thousands of genes. In these data sets, the batch
effects are particularly strong as the data comes from different institutions
and was obtained using different experimental protocols. The visualizations
constructed by our proposed approach are cleared of batch effects, and the cells
from secondary data sets correctly co-cluster with cells from the primary data
sharing the same cell type.

@@

Two-dimensional embeddings and their visualizations may assist in the analysis
and interpretation of high-dimensional data. Intuitively, two data instances
should be co-located in the resulting visualization if their multi-dimensional
profiles are similar. For this task, non-linear embedding techniques such as
t\nobreakdash -distributed stochastic neighbor embedding (t-SNE)~\cite{tsne} or uniform
manifold approximation and projection~\cite{umap} have recently complemented
traditional data transformation and embedding approaches such as principal
component analysis (PCA) and multi-dimensional
scaling~\cite{distill,umap_single_cell}. While useful for visualizing data
from a single coherent source, these methods may encounter problems if the data
comes from multiple sources. Here, when performing dimensionality reduction on a merged data set, the resulting visualizations would typically
reveal source-specific clusters instead of grouping data instances of the same class-type, regardless of data sources. This source-specific confounding is often referred to as {\em domain
shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift} or
{\em dataset shift}~\cite{dataset_shift}. In bioinformatics, the
domain-specific differences are more commonly referred to as {\em batch
effects}~\cite{cca,mnn,seurat}.

Massive, multi-variate biological data sets often suffer from these source-specific
biases. Consider an example from single-cell genomics, the domain we will focus
on in this manuscript and that was --- besides current scientific challenges
--- selected also due to the availability and abundance of recently published
data. Single-cell RNA sequencing (scRNA-seq) data sets are the result of isolating RNA molecules from individual cells,
which serve as an estimate of the expression of cell's genes.
Single-cell studies, which can exceed thousands of cells and tens of thousands of genes,
typically start with the analysis of cell types. Here, it is generally expected that
cells of the same type would cluster together in two-dimensional data
visualisation~\cite{seurat}. For instance, Fig.~\ref{fig:batch_effect}.a shows
t-SNE embedded data from mouse brain cells originating from the visual
cortex~\cite{hrvatin2018} and the hypothalamus~\cite{chen2017}. The figure reveals
distinct clusters but also separates the data from the two brain regions. These
two regions share the same cell types and --- contrary to the depiction in
Fig.~\ref{fig:batch_effect}.a --- we would expect the data points from the two
studies to overlap. Batch effects similarly prohibit the utility of t-SNE in
the exploration of pancreatic cells in Fig.~\ref{fig:batch_effect}.b, which
renders the data from a pancreatic cell atlas~\cite{baron2016} and similarly-typed
cells from diabetic patients~\cite{xin2016}. Just like with data from brain
cells, pancreatic cells cluster primarily by data source, again
resulting in an uninformative visualization driven by batch
effect.

@@

Current solutions to embedding the data from various data sources address the
batch effect problems up-front. The data is typically preprocessed and
transformed such that the batch effects are explicitly
removed. Recently proposed procedures for batch effect removal include
canonical correlation analysis~\cite{cca} and mutual
nearest-neighbors~\cite{mnn,seurat}.
In these works, batch effects are deemed removed when cells from different sources exhibit good mixing in a t-SNE visualization.
The elimination of batch effects may require aggressive data
preprocessing which may blur the boundaries between cell types. Another problem
is also the inclusion of any new data, for which the entire data analysis
pipeline must be rerun, usually resulting in a different layout and clusters
that have little resemblance to original visualization and thus require
reinterpretation.

We propose a direct solution of rendering t-SNE visualizations that addresses
batch effects. Our approach treats one of the data sets as a {\em reference}
and aims to embed the cells from another, {\em secondary data set} to a common low-dimensional space.
We construct a t\nobreakdash -SNE embedding using the reference data set, and then use it
as a scaffold for the embedding of data points from the secondary data.
The key idea underpinning our approach is that data points are embedded
independently of one another.

Independence of each new
embedded data instance from the secondary data set causes the clustering landscape
to depend only on the reference scaffold, thus removing data source-driven
variation. In other words, when including new data, the scaffold inferred from
the reference data set is kept unchanged and defines a ``gravitational
field'', independently driving the embedding of each new instance. For example, in
Fig.~\ref{fig:transform_brain}, the cells from the visual cortex define the
scaffold (Fig.~\ref{fig:transform_brain}.a) into which we embed the cells from the
hypothalamus (Fig.~\ref{fig:transform_brain}.b). Unlike in their joint t\nobreakdash -SNE
visualization (Fig.~\ref{fig:batch_effect}.a), the hypothalamic cells are
dispersed across the entire embedding space and their cell type correctly matches
the prevailing type in reference clusters.

@@

t-SNE is a local, non-linear dimensionality reduction method, tailored to the visualisation of
high-dimensional data sets. Given a multi-dimensional data set $\mathbf{X} =
\left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in
\mathbb{R}^D$ where $N$ is the number of samples in the reference data set,
t-SNE aims to find a low dimensional embedding $\mathbf{Y} = \left \{
\mathbf{y}_1, \mathbf{y}_2, \dots, \mathbf{y}_N \right \} \\\in \mathbb{R}^d$
where $d \ll D$, such that if points $\mathbf{x}_i$ and $\mathbf{x}_j$ are
close in the multi-dimensional space, their corresponding embeddings
$\mathbf{y}_i$ and $\mathbf{y}_j$ are also close. Since t-SNE is primarily used
as a visualization tool, $d$ is typically set to two. The similarity between
two data points in t-SNE is defined as:

@@

The time complexity needed to evaluate the similarities in
Eq.~\ref{eq:cauchy_kernel} is $\mathcal{O}(N^2)$, making the application of
the algorithm impractical for any reasonably-sized data. To address larger data
sets, we adopt a recent approach for low-rank approximation of
gradients based on polynomial interpolation which reduces the time
complexity of t-SNE to $\mathcal{O}(N)$. This approximation enables the
visualization of massive data sets, possibly containing millions of data
points~\cite{fi_tsne}.

The resulting embeddings substantially depend on the value of the perplexity
parameter. Perplexity can be interpreted as the number of
neighbors for which the distances in the embedding space are preserved. Small
values of perplexity result in tightly-packed clusters of points and effectively
ignore the long-range interactions between clusters. Larger values may result
in a more globally consistent visualisations, preserving distances on a large
scale and organizing clusters in a more meaningful way. Larger values of
perplexity can lead to merging of multiple small clusters, thus obscuring
local aspects of the data~\cite{art_of_using_tsne}.

@@

In the optimization step, we refine point positions using batch gradient
descent. We use an adaptive learning rate scheme with momentum 
to speed up the convergence, as proposed by Jacobs~\cite{momentum,bh_tsne}. We run gradient descent
with momentum $\alpha$ set to $0.8$ for $250$ iterations, where the
optimization converged in all our experiments. The time complexity
needed to evaluate the gradients in Eq.~\ref{eq:gradient} is $\mathcal{O}(N
\cdot M)$, however, by adapting the same polynomial interpolation based
approximation, this is reduced to $\mathcal{O}(N)$.

@@

We apply the proposed approach to t-SNE visualizations of
single-cell data. Data in this realm include a variety of
cells from specific tissues, and characterize them through gene expression.
In our experiments, we considered several recently published data
sets where cells were annotated with the cell type. Our aim was to construct
t-SNE visualizations where similarly-typed cells would cluster together,
despite systematic differences between data sources. To that end, we focus
on comparing different ways of using t-SNE rather than differences to embeddings like PCA or MDS, which
have been substantially covered before~\todo{Add citation}. 
Below, we list the data sets,
describe single-cell specific data preprocessing procedures, and display the
resulting data visualizations. Finally, we discuss the success of the proposed
approach in alleviating the batch effects.

@@

We use three pairs of reference and secondary single-cell data sets originating
from different organisms and tissues. The data in each pair were chosen so that
the majority of cell types from the secondary data set were included in the
reference set (Table~\ref{tab:datasets}).

\begin{table}[ht]
\begin{center}
\setlength\tabcolsep{4pt}
\begin{tabular}{l c c r c c}
\toprule
Study & Organism/Tissue & Protocol & Cells & Cell Types & Sparsity (\%) \\
\midrule
Hrvatin \etal & \multirow{2}{*}{mouse brain} & inDrop & 48,266 & 9 & 94 \\
Chen \etal & & Drop-seq & 14,437 & 6 & 93 \\[5pt]
Baron \etal & \multirow{2}{*}{human pancreas} & inDrop & 8,569 & 9 & 91 \\
Xin \etal & & SMARTer & 1,492 & 4 & 86 \\[5pt]
Macosko \etal & \multirow{2}{*}{mouse retina} & Drop-seq & 44,808 & 12 & 97 \\
Shekhar \etal & & Drop-seq & 27,499 & 5 & 96 \\
\bottomrule
\end{tabular}
\end{center}
\caption{Data sets used in our experiments. In each pair, the first data set
(Hrvatin \etal, Baron \etal, and Macoscko \etal) was used as a reference. In
all cases, we relied on the quality control and annotations from the original
publication. To facilitate comparisons, the cell annotations were harmonized using
cell type annotations from the cell ontology~\cite{cell_ontology}. Notice that
different RNA sequencing protocols were used to estimate gene expressions. We
report the number of cell types from each data set retained after
preprocessing. Single-cell data is sparse, typically containing less than 10\%
expressed genes per cell.}
\label{tab:datasets}
\end{table}

The cells in the data sets originates from the following three tissues:
\begin{description}
\item[Mouse brain.] The data set from Hrvatin \etal~\cite{hrvatin2018} contains
cells from the visual cortex exploring transcriptional changes after exposure
to light. This was used as a reference for the data from Chen
\etal~\cite{chen2017}, containing various cells from the mouse hypothalamus and
their reaction to food deprivation. From the secondary data, we removed cells
with no corresponding types in the reference, namely ependymal cells,
epithelial cells, tanycytes, and unlabelled cells.

@@

\item[Mouse retina.] The data set from Macosko \etal~\cite{macosko2015} was
created as an atlas of mouse retinal cell types. We used this as a reference
for the data from Shekhar \etal~\cite{shekhar2016}, who built an atlas for
different types of retinal bipolar cells.
\end{description}

@@

Due to the specific nature of single-cell data, additional steps must be taken
to properly apply t-SNE. We use a standard single-cell preprocessing pipeline,
consisting of the selection of 3,000 representative genes (see
Sec.~\ref{sec:gene-selection}), library size normalization, log-transformation,
standardization, and PCA-based representation that retains 50 principal
components~\cite{seurat,scanpy}. To obtain the reference embedding, we apply
multi-scale t-SNE using PCA initialization~\cite{art_of_using_tsne}. Due to
high-dimensionality of the preprocessed input data we use cosine distance to
estimate similarities between reference data points~\cite{Domingos2012-CACM}.
When adding new data points from the secondary data set to the reference embedding,
we select 1,000 genes present in both data sets and use these to estimate the
similarities between the secondary data item and reference data points. The
similarities are estimated using cosine similarity. We note that similarities are
computed using the raw count matrices. The preprocessing stages are detailed in
accompanying Python notebooks (Sec.~\ref{sec:implementation}).

@@

Additional experiments with embedding secondary datasets from independent
studies are provided in the accompanying Python notebooks. We embedded brain
cells from \cite{} to the reference embedding shown in
Fig.~\ref{fig:transform_brain}.a, and pancreas cells from~\cite{} to reference
in \ref{fig:transform_pancreas}.a. The conclusions are largely consistent, with the error rates 
of X\% and Y\%. \todo{Add errors}

@@

In principle, our theoretically-grounded embedding of secondary data into the
scaffold defined by the reference embedding could be simplified with the
application of the nearest neighbors-based procedure. For example, while
describing a set of tricks for t-SNE, Kobak and Berens~\cite{art_of_using_tsne}
proposed positioning new points into a known embedding by placing them in the
median position of their 10 nearest neighbors, where the neighborhood was
estimated in the original data space. Notice that we use this approach as well,
but only for the initialization of positions of new data instances that are
subject to further optimization. Despite both nearest-neighbours search and
t-SNE optimization can be computed in linear time, the former dominates the
runtime (mouse retina example; M reference, N secondary cells, X min N-search,
Y min optimization).

We demonstrate a case where nearest neighbor-based positioning is insufficient
in Fig.~\ref{fig:optimization}. It may yield clumped visualizations where the
optimal positioning using the t-SNE loss function is much more dispersed and
rightfully shows a more considerable variation in the secondary data. 
Some data points may also fall into the empty regions
between different clusters, while after optimization they typically
move closer to same-typed groups. While the nearest neighbour-based approach
always places new data points to their most similar cluster, our proposed
t-SNE approach can identify extreme outliers and place them far from the
reference cluster.

@@

Our procedure and the resulting cell landscape are affectected by the asymmetry
of determining the reference and the secondary datasets. In practice, however,
newly produced \em secondary \em data would be embedded into published and
annotated reference landscapes. Scaling-up to a wider collection of cell
types is driven by increasingly large collections of data, e.g the Human Cell
Atlas initiative~\cite{hca}.

@@

Almost all recent publications of single-cell studies begin with a
two-dimensional visualization of the data that reveals the cellular diversity containing many 
different cell-types from the study. While any dimensionality reduction
technique can be used to render such a visualization, different variants of
t-SNE are most often used. Due to the ability to explore biological mechanisms
at the individual cell level, single-cell studies are increasingly widespread, and
their publications in the past couple of years are abundant. One of the central
tasks in single-cell studies is the classification of new cells based on
findings from previous studies. Such transfer of knowledge is often difficult
due to batch effects present in data from different sources.
Addressing batch effects by adapting and extending t-SNE, the prevailing method used
to present single-cell data in two-dimensional visualization motivated the research
presented in this paper.

>>>>

Dimensionality reduction techniques, such as t-SNE, can construct informative
visualizations of high-dimensional data. When working with multiple data sets,
a straightforward application of these methods often fails; instead of
revealing underlying classes, the resulting visualizations expose
data set-specific clusters. To circumvent these batch effects, we propose an
embedding procedure that takes a t-SNE visualization constructed on a reference
data set and uses it as a scaffold for embedding new data. The new, secondary data is
embedded one data point at the time. This prevents any interactions
between instances in the secondary data and implicitly mitigates batch
effects. We demonstrate the utility of this approach by analyzing six
recently published single-cell gene expression data sets with up to tens
of thousands of cells and thousands of genes. The batch
effects are particularly strong as the used data comes from different institutions
and was obtained using different experimental protocols. The visualizations
constructed by our proposed approach are cleared of batch effects, and the cells
from secondary data sets correctly co-cluster with cells of the same type from the primary data.

@@

Two-dimensional embeddings and their visualizations may assist in the analysis
and interpretation of high-dimensional data. Intuitively, two data instances
should be co-located in the resulting visualization if their multi-dimensional
profiles are similar. For this task, non-linear embedding techniques such as
t\nobreakdash -distributed stochastic neighbor embedding (t-SNE)~\cite{tsne} or uniform
manifold approximation and projection~\cite{umap} have recently complemented
traditional data transformation and embedding approaches such as principal
component analysis (PCA) and multi-dimensional
scaling~\cite{distill,umap_single_cell}. While useful for visualizing data
from a single coherent source, these methods may encounter problems 
with multiple data sources. Here, when performing dimensionality reduction on a merged data set, the resulting visualizations would typically
reveal source-specific clusters instead of grouping data instances of the same class, regardless of data sources. This source-specific confounding is often referred to as {\em domain
shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift} or
{\em data set shift}~\cite{data set_shift}. In bioinformatics, the
domain-specific differences are more commonly referred to as {\em batch
effects}~\cite{cca,mnn,seurat}.

Massive, multi-variate biological data sets often suffer from these
source-specific biases.  The focus of this work is single-cell genomics, a
domain that was selected due high biomedical relevance and and abundance of
recently published data. Single-cell RNA sequencing (scRNA-seq) data sets are
the result of isolating RNA molecules from individual cells,
which serve as an estimate of the expression of cell's genes.
The studies can exceed thousands of cells and tens of thousands of genes,
and typically start with cell type analysis. Here, it is expected that
cells of the same type would cluster together in two-dimensional data
visualisation~\cite{seurat}. For instance, Fig.~\ref{fig:batch_effect}.a shows
t-SNE embedded data from mouse brain cells originating from the visual
cortex~\cite{hrvatin2018} and the hypothalamus~\cite{chen2017}. The figure reveals
distinct clusters but also separates the data from the two brain regions. These
two regions share the same cell types and --- contrary to the depiction in
Fig.~\ref{fig:batch_effect}.a --- we would expect the data points from the two
studies to overlap. Batch effects similarly prohibit the utility of t-SNE in
the exploration of pancreatic cells in Fig.~\ref{fig:batch_effect}.b, which
renders the data from a pancreatic cell atlas~\cite{baron2016} and similarly-typed
cells from diabetic patients~\cite{xin2016}. Just like with data from brain
cells, pancreatic cells cluster primarily by data source, again
resulting in a visualization driven by batch effects.

@@

Current solutions to embedding the data from various data sources address the
batch effect problems up-front. The data is typically preprocessed and
transformed such that the batch effects are explicitly
removed. Recently proposed procedures for batch effect removal include
canonical correlation analysis~\cite{cca} and mutual
nearest-neighbors~\cite{mnn,seurat}.
In these works, batch effects are deemed removed when cells from different sources exhibit good mixing in a t-SNE visualization.
The elimination of batch effects may require aggressive data
preprocessing which may blur the boundaries between cell types. Another problem
is also the inclusion of any new data, for which the entire analysis
pipeline must be rerun, usually resulting in a different layout and clusters
that have little resemblance to original visualization and thus require
reinterpretation.

We propose a direct solution of rendering t-SNE visualizations that addresses
batch effects. Our approach treats one of the data sets as a {\em reference}
and aims to embed the cells from another, {\em secondary data set} to a common low-dimensional space.
We construct a t\nobreakdash -SNE embedding using the reference data set, and then use it
as a scaffold for the embedding of the secondary data.
The key idea underpinning our approach is that secondary data points are embedded
independently of one another.

Independent embedding of each secondary datum causes the clustering landscape
to depend only on the reference scaffold, thus removing data source-driven
variation. In other words, when including new data, the scaffold inferred from
the reference data set is kept unchanged and defines a ``gravitational
field'', independently driving the embedding of each new instance. For example, in
Fig.~\ref{fig:transform_brain}, the cells from the visual cortex define the
scaffold (Fig.~\ref{fig:transform_brain}.a) into which we embed the cells from the
hypothalamus (Fig.~\ref{fig:transform_brain}.b). Unlike in their joint t\nobreakdash -SNE
visualization (Fig.~\ref{fig:batch_effect}.a), the hypothalamic cells are
dispersed across the entire embedding space and their cell type correctly matches
the prevailing type in reference clusters.

@@

Local, non-linear dimensionality reduction by t-SNE is performed as follows.
Given a multi-dimensional data set $\mathbf{X} =
\left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in
\mathbb{R}^D$ where $N$ is the number of data points in the reference data set,
t-SNE aims to find a low dimensional embedding $\mathbf{Y} = \left \{
\mathbf{y}_1, \mathbf{y}_2, \dots, \mathbf{y}_N \right \} \\\in \mathbb{R}^d$
where $d \ll D$, such that if points $\mathbf{x}_i$ and $\mathbf{x}_j$ are
close in the multi-dimensional space, their corresponding embeddings
$\mathbf{y}_i$ and $\mathbf{y}_j$ are also close. Since t-SNE is primarily used
as a visualization tool, $d$ is typically set to two. The similarity between
two data points in t-SNE is defined as:

@@

The time complexity needed to evaluate the similarities in
Eq.~\ref{eq:cauchy_kernel} is $\mathcal{O}(N^2)$, making the application of
the algorithm impractical for large data sets. 
Hence, we adopt a recent approach for low-rank approximation of
gradients based on polynomial interpolation which reduces the time
complexity of t-SNE to $\mathcal{O}(N)$. This approximation enables the
visualization of massive data sets, possibly containing millions of data
points~\cite{fi_tsne}.

The resulting embeddings substantially depend on the value of the perplexity
parameter. Perplexity can be interpreted as the number of
neighbors for which the distances in the embedding space are preserved. Small
values of perplexity result in tightly-packed clusters of points and effectively
ignore the long-range interactions between clusters. Larger values may result
in a more globally consistent visualisations --- preserving distances on a large
scale and organizing clusters in a more meaningful way --- but can lead to merging 
small clusters and thus obscuring local aspects of the data~\cite{art_of_using_tsne}.

@@

In the optimization step, we refine point positions using batch gradient
descent. We use an adaptive learning rate scheme with momentum 
to speed up the convergence, as proposed by Jacobs~\cite{momentum,bh_tsne}. We run gradient descent
with momentum $\alpha$ of $0.8$ for $250$ iterations, where the
optimization converged in all our experiments. The time complexity
needed to evaluate the gradients in Eq.~\ref{eq:gradient} is $\mathcal{O}(N
\cdot M)$, however, by adapting the same polynomial interpolation based
approximation, this is reduced to $\mathcal{O}(N)$.

@@

We apply the proposed approach to t-SNE visualizations of
single-cell data. Data in this realm include a variety of
cells from specific tissues and are characterized through gene expression.
In our experiments, we considered several recently published data
sets where cells were annotated with the cell type. Our aim was to construct
t-SNE visualizations where similarly-typed cells would cluster together,
despite systematic differences between data sources. To that end, we focus
on comparing different ways of using t-SNE rather than differences to embeddings like PCA or MDS, which
have been substantially covered before~\todo{Add citation}. 
Below, we list the data sets,
describe single-cell specific data preprocessing procedures, and display the
resulting data visualizations. Finally, we discuss the success of the proposed
approach in alleviating the batch effects.

@@

We use three pairs of reference and secondary single-cell data sets originating
from different organisms and tissues. The data in each pair were chosen so that
the majority of cell types from the secondary data set were included in the
reference set (Table~\ref{tab:data sets}).
The cells in the data sets originates from the following three tissues:

\begin{description}
\item[Mouse brain.] The data set from Hrvatin \etal~\cite{hrvatin2018} contains
cells from the visual cortex exploring transcriptional changes after exposure
to light. This was used as a reference for the data from Chen
\etal~\cite{chen2017}, containing cells from the mouse hypothalamus and
their reaction to food deprivation. From the secondary data, we removed cells
with no corresponding types in the reference: tanycytes, ependymal,
epithelial, and unlabelled cells.

@@

\item[Mouse retina.] The data set from Macosko \etal~\cite{macosko2015} was
created as an atlas of mouse retinal cell types. We used this as a reference
for the data from Shekhar \etal~\cite{shekhar2016}, who built an atlas for retinal bipolar cells.
\end{description}

\vspace{-3mm}
\begin{table}[ht]
\begin{center}
\setlength\tabcolsep{4pt}
\begin{tabular}{l c c r c c}
\toprule
Study & Organism/Tissue & Protocol & Cells & Cell Types & Sparsity (\%) \\
\midrule
Hrvatin \etal & \multirow{2}{*}{mouse brain} & inDrop & 48,266 & 9 & 94 \\
Chen \etal & & Drop-seq & 14,437 & 6 & 93 \\[5pt]
Baron \etal & \multirow{2}{*}{human pancreas} & inDrop & 8,569 & 9 & 91 \\
Xin \etal & & SMARTer & 1,492 & 4 & 86 \\[5pt]
Macosko \etal & \multirow{2}{*}{mouse retina} & Drop-seq & 44,808 & 12 & 97 \\
Shekhar \etal & & Drop-seq & 27,499 & 5 & 96 \\
\bottomrule
\end{tabular}
\end{center}
\caption{Data sets used in our experiments. The first data set in a pair
(Hrvatin \etal, Baron \etal, and Macoscko \etal) was used as a reference. 
We relied on the quality control and annotations from the original
publication and report the humber of cell types after preprocessinng. 
The cell annotations were made consistent to
annotations from the Cell Ontology~\cite{cell_ontology}. Notice that
different RNA sequencing protocols were used to estimate gene expressions.}
\label{tab:data sets}
\end{table}

@@

Due to the specific nature of single-cell data, additional steps must be taken
to properly apply t-SNE. We use a standard single-cell preprocessing pipeline,
consisting of the selection of 3,000 representative genes (see
Sec.~\ref{sec:gene-selection}), library size normalization, log-transformation,
standardization, and PCA-based representation that retains 50 principal
components~\cite{seurat,scanpy}. To obtain the reference embedding, we apply
multi-scale t-SNE using PCA initialization~\cite{art_of_using_tsne}. Due to
high-dimensionality of the preprocessed input data we use cosine distance to
estimate similarities between reference data points~\cite{Domingos2012-CACM}.
When adding new data points from the secondary data set to the reference embedding,
we select 1,000 genes present in both data sets and use the cosine similarity 
to estimate the similarities between the secondary data item and reference data
points.  We note that similarities are computed using the raw count matrices.
The preprocessing stages are detailed in accompanying Python notebooks
(Sec.~\ref{sec:implementation}).

@@

Additional experiments with embedding secondary data sets from independent
studies are provided in the accompanying Python notebooks. We embedded brain
cells from \cite{} to the reference embedding shown in
Fig.~\ref{fig:transform_brain}.a, and pancreas cells from~\cite{} to reference
in \ref{fig:transform_pancreas}.a. The conclusions are largely consistent, with the error rates 
of X\% and Y\%. \todo{Add errors}

@@

In principle, our theoretically-grounded embedding of secondary data into the
scaffold defined by the reference embedding could be simplified with the
application of the nearest neighbors-based procedure. For example, while
describing a set of tricks for t-SNE, Kobak and Berens~\cite{art_of_using_tsne}
proposed positioning new points into a known embedding by placing them in the
median position of their 10 nearest neighbors, where the neighborhood was
estimated in the original data space. Notice that we use this approach as well,
but only for the initialization of positions of new data instances that are
subject to further optimization. Despite both nearest-neighbours search and
t-SNE optimization can be computed in linear time, the former dominates the
runtime (mouse retina example; M reference, N secondary cells, X min N-search,
Y min optimization). \todo{Add nunmbers}

We demonstrate a case where nearest neighbor-based positioning is insufficient
in Fig.~\ref{fig:optimization}. It may yield clumped visualizations where the
optimal positioning using the t-SNE loss function is much more dispersed and
rightfully shows a more considerable variation in the secondary data. 
Some points may also fall into the empty regions, while after optimization they typically
move closer to same-typed groups. While the nearest neighbour-based approach
always places new data points to their most similar cluster, our proposed
t-SNE approach can identify extreme outliers and place them far from the
reference cluster.

@@

Our procedure and the resulting cell landscape are therefore affectected by the asymmetry
of determining the reference and the secondary data sets. In practice, however,
newly produced \em secondary \em data would be embedded into published and
annotated reference landscapes. Scaling-up to a wider collection of cell
types is driven by increasingly large collections of data, e.g the Human Cell
Atlas initiative~\cite{hca}.

@@

Almost all recent publications of single-cell studies begin with a
two-dimensional visualization of the data that reveals the cellular diversity. 
While most dimensionality reduction
techniques can be used to render such a visualization, different variants of
t-SNE are most often used. Due to the ability to explore biological mechanisms
at the individual cell level, single-cell studies are increasingly widespread, and
their publications in the past couple of years are abundant. One of the central
tasks in single-cell studies is the classification of new cells based on
findings from previous studies. Such transfer of knowledge is often difficult
due to batch effects present in data from different sources.
Addressing batch effects by adapting and extending t-SNE, the prevailing method used
to present single-cell data in two-dimensional visualization motivated the research
presented in this paper.
